# Reproducible Research: Peer Assessment 1

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(ggplot2)
library(scales)
library(Hmisc)
```

## Loading and preprocessing the data
##### 1. Load the data (i.e. read.csv())
```{r, results='markup', warning=TRUE, message=TRUE}
if(!file.exists("activity.csv")) {
        tempfile <- tempfile()
        download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",destfile = tempfile)
        unzip(tempfile)
        unlink(tempfile)
}
#load data
activity_data <- read.csv("activity.csv")
```
######Inspecting the data sets:
```{r}
summary(activity_data)
```
##### 2. Process/transform the data (if necessary) into a format suitable for your analysis
```{r}
activity_data$date <- as.POSIXct(activity_data$date, format="%Y-%m-%d")

# Compute the weekdays from the date attribute
activity_data <- data.frame(date=activity_data$date, 
                           weekday=tolower(weekdays(activity_data$date)), 
                           steps=activity_data$steps, 
                           interval=activity_data$interval)

# Compute the day type (weekend or weekday)
activity_data <- cbind(activity_data, 
                      daytype=ifelse(activity_data$weekday == "saturday" | 
                                     activity_data$weekday == "sunday", "weekend", 
                                     "weekday"))

# Create the final data.frame
activity <- data.frame(date=activity_data$date, 
                       weekday=activity_data$weekday, 
                       daytype=activity_data$daytype, 
                       interval=activity_data$interval,
                       steps=activity_data$steps)
```

#####Checking the transformed data
```{r}
summary(activity)
```

-----

## What is mean total number of steps taken per day?
```{r}
stepsByDay <- tapply(activity$steps, activity$date, sum, na.rm=TRUE)

```

##### 1. Make a histogram of the total number of steps taken each day
```{r}
hist(stepsByDay, xlab = "Steps per Day", main = "Total number of steps taken per day", col = "wheat")
```

##### 2. Calculate and report the mean and median total number of steps taken per day
```{r}
stepsByDayMean <- mean(stepsByDay)
stepsByDayMedian <- median(stepsByDay)
```
* The Mean is `r stepsByDayMean`
* The Median is `r stepsByDayMedian`

-----

## What is the average daily activity pattern?
```{r}
average_steps_time_block <- aggregate(x=list(meanSteps=activity_data$steps), by=list(interval=activity_data$interval), FUN=mean, na.rm=TRUE)
```

##### 1. Make a time series plot
```{r}
ggplot(data=average_steps_time_block, aes(x=interval, y=meanSteps)) +
    geom_line() +
    xlab("5-Minute Interval") +
    ylab("Average Number of Steps") 
```

##### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
mostSteps <- which.max(average_steps_time_block$meanSteps)
timeMostSteps <-  gsub("([0-9]{1,2})([0-9]{2})", "\\1:\\2", average_steps_time_block[mostSteps,'interval'])
```

* The 5-minute interval with the most steps is `r timeMostSteps`

----

## Imputing missing values
##### 1. Calculate and report the total number of missing values in the dataset 
```{r}
numMissingValues <- length(which(is.na(activity_data$steps)))
```

* The Number of missing values is `r numMissingValues`

##### 2. Devise a strategy for filling in all of the missing values in the dataset.
```{r}
imputedactivitydata <- activity_data
imputedactivitydata$steps[is.na(imputedactivitydata$steps)] <- mean(imputedactivitydata$steps,na.rm=TRUE)
imputedactivitydata$steps <- as.numeric(imputedactivitydata$steps)
imputedactivitydata$interval <- as.numeric(imputedactivitydata$interval)
colSums(is.na(imputedactivitydata))
```
##### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
summary(imputedactivitydata)
```


##### 4. Make a histogram of the total number of steps taken each day 
```{r}
stepsByDayImputed <- tapply(imputedactivitydata$steps, imputedactivitydata$date, sum)
qplot(stepsByDayImputed, xlab='Total steps per day (Imputed)', ylab='Frequency using binwith 2500', binwidth=2500)
```

##### ... and Calculate and report the mean and median total number of steps taken per day. 
```{r}
stepsByDayMeanImputed <- as.integer(mean(stepsByDayImputed))
stepsByDayMedianImputed <- as.integer(median(stepsByDayImputed))
```
* The Imputed Mean and Median are `r stepsByDayMeanImputed` and `r stepsByDayMedianImputed` resoectively.


----

## Are there differences in activity patterns between weekdays and weekends?
##### 1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
imputedactivitydata$dateType <-  ifelse(as.POSIXlt(imputedactivitydata$date)$wday %in% c(0,6), 'weekend', 'weekday')
```

##### 2. Make a panel plot containing a time series plot

```{r}
averagedactivity <- aggregate(steps ~ interval + dateType, data=imputedactivitydata, mean)
ggplot(averagedactivity, aes(interval, steps)) + 
    geom_line() + 
    facet_grid(dateType ~ .) +
    xlab("5-minute interval") + 
    ylab("avarage number of steps")
```
